Life in the canopy: community trait assessments reveal substantial functional diversity among fern epiphytes

### 2019-10-28 ms/SI.rmd
Update MS
final changes before initial sumbission to New Phytologist

ggsave(plot = fig_s1, filename = here::here("ms/Fig_S1.pdf"), height = 8, width = 4, units = "in")
```

@@

```{r fig-s1}
loadd(cwsd_scatterplot, cache = canopy_cache)
ggsave(plot = cwsd_scatterplot, filename = here::here("ms/Fig_S2.pdf"), height = 6, width = 7, units = "in")
```

>>>>

ggsave(plot = fig_s1, filename = here::here("ms/Fig_S1.eps"), height = 8, width = 4, units = "in")
```

@@

```{r fig-s1}
loadd(cwsd_scatterplot, cache = canopy_cache)
ggsave(plot = cwsd_scatterplot, filename = here::here("ms/Fig_S2.eps"), height = 6, width = 7, units = "in")
```

### 2019-10-28 ms/manuscript.Rmd
Update MS
final changes before initial sumbission to New Phytologist

```{r word-count table}
tibble(
  words = c(
    "Total word count (excluding summary, references and legends):_n",
    "Summary:_n",
    "Introduction:_n",
    "Materials and Methods:_n",
    "Results:_n",
    "Discussion:_n",
    "Acknowledgements:_n"),
  figs_tabs = c(
    "No. of figures:_7 (all in color)",
    "No. of tables:_5",
    "No. of supporting information files:_7 (Fig. S1; Table S1-S6)", "", "", "", "")
) %>%
  separate(words, c("words", "count"), sep = "_") %>%
  separate(figs_tabs, c("figs_tabs", "n"), sep = "_", fill = "right") %>%
  mutate(n = replace_na("")) %>%
  gt() %>%
  cols_label(
    words = "",
    count = "",
    figs_tabs = "",
    n = "",
  ) %>% 
  as_rtf() %>%
  write_lines(here::here("ms/word_count.rtf"))
```

@@

**Key words:** community ecology, functional traits, gametophyte, Moorea, Society Islands, Tahiti.

@@

While many similar adaptations are found in epiphytic seed plants and serve as key reference points [@Benzing1990], the evolution of epiphytism in ferns is more complex owing to their unique life-cycle. 
Ferns produce a free-living gametophyte that is typically smaller than the sporophyte and relies on aquatic motile sperm.
Unlike the sporophyte, the fern gametophyte lacks vascular tissue, stomata, or cuticle, placing it in constant equilibrium with its external environment. 
In parallel with some of the adaptations in the sporophytes, the gametophytes of some epiphytic ferns also exhibit morphological and physiological adaptations to minimize water loss and tolerate water scarcity. 
For example, extreme desiccation tolerance is prevalent in gametophytes of multiple lineages of epiphytic ferns, but is rare in sporophytes [@WatkinsJr.2007c; @Pittermann2013].
In addition, epiphytic fern gametophytes tend to exhibit more complex three-dimensional morphologies that may function to retain water and slow its loss [@WatkinsJr.2007c].
Despite evidence for adaptations to epiphytism in multiple lineages in both sporophytes and gametophytes, no previous studies to our knowledge have analyzed the functional ecology of fern epiphytes in a phylogenetic context including both life stages and representatives of all five Cretaceous epiphytic radiations *sensu* [@Schuettpelz2009].

@@

Here, we seek to understand the role of the canopy in fern diversification by investigating traits related to epiphytic growth in fern gametophytes and sporophytes on Moorea, French Polynesia.
Moorea is an isolated tropical oceanic island with high phylogenetic diversity of ferns distributed across a steep elevational gradient, making it ideal for a community phylogenetic analysis of functional traits. 
Specifically, we test the following hypotheses related to epiphytic growth in ferns:

1. Epiphytic ferns have converged upon similar morphologies in both sporophyte and gametophyte phases.

2. Epiphytic fern communities have lower functional and phylogenetic diversity relative to terrestrial communities.

@@

Terrestrial and epiphytic species are partly differentiated in trait space, occupying mostly distinct areas but overlapping in the middle regardless of method used (standard or phylogenetic PCA) (\fig \@ref(fig:pca)(c, d)). The first two PC axes explained `r percent(pca_std_cum_var)` and `r percent(pca_phy_cum_var)` of the variation in the trait data when analyzed using standard and phylogenetic PCA, respectively. In the standard PCA, size-related traits tended to load on PC1, with SLA, number of pinnae pairs, and degree of dissection on PC2. In the phylogenetic PCA, SLA was strongly loaded on PC1, while dissection, pinnae, and other size-related traits were mostly loaded on PC2.

@@

In the full-subsets analysis including all climate variables and diversity metrics, growth habit and temperature emerged as the most important variables (\fig \ref{fig:heatmap}). Most of the the best-fitting models included growth habit or the interaction of growth habit and a temperature-related variable (\tab S6). Spatial autocorrelation was not detected for any of the diversity metrics (\tab S6).

@@

Despite the apparent constraints on size and the recurrence of non-cordate gametophytes exhibited by epiphytes, we observed greater morphological diversity among epiphytes versus terrestrial species for several traits.
These results do not support the hypothesis that fern epiphytes have reduced functional diversity.
For example, although epiphytes demonstrated less dissected fronds at the species level (\tab \@ref(tab:pics)), they exhibited much greater variation in this trait relative to terrestrial ferns at the community level (\fig S2).
It is likely that the high diversity of this trait reflects different physiological adaptations to various canopy niches. 
For instance, filmy ferns (Hymenophyllaceae), which depend on desiccation tolerance to survive dry conditions, have extremely thin, highly dissected fronds.
Bird’s nest ferns (e.g., *Asplenium nidus* L.), in contrast, avoid desiccation by capturing detritus and water in the bases of their simple, rosette leaves. 
At low-elevation sites on Moorea where these ferns co-occur, filmy ferns frequently grow in more humid portions of trees including near the main trunk, whereas bird’s nest ferns often grow higher in the canopy and on more exposed, drought-prone branches (J. Nitta, pers. obs.). 
Similar predictable patterns of niche partitioning at the local scale has been reported in neotropical epiphtyic ferns as well [@Hietz1998; @Tausz2001]. 
This distribution may reflect alternative water-usage strategies aligned to specific microhabitats in the canopy [desiccation tolerance vs. water capture; @Hietz1998]. 
Such hypotheses need to be carefully tested with additional quantitative investigation, including more finely examining microclimatic variation and species composition across the canopy.

@@

### The primacy of growth habit in determining fern community composition as indicated by modelling

>>>>

```{r word-count table}
tibble(
  words = c(
    "Total word count (Introduction, Materials and Methods, Results, Discussion):_n",
    "Summary:_n",
    "Introduction:_n",
    "Materials and Methods:_n",
    "Results:_n",
    "Discussion:_n",
    "Acknowledgements:_n"),
  figs_tabs = c(
    "No. of figures:_7 (all in color)",
    "No. of tables:_5",
    "No. of supporting information files:_8 (Figs. S1-S2; Tables S1-S6)", "", "", "", "")
) %>%
  separate(words, c("words", "count"), sep = "_") %>%
  separate(figs_tabs, c("figs_tabs", "n"), sep = "_", fill = "right") %>%
  mutate(n = replace_na("")) %>%
  gt() %>%
  cols_label(
    words = "",
    count = "",
    figs_tabs = "",
    n = "",
  ) %>% 
  as_rtf() %>%
  write_lines(here::here("ms/word_count.rtf"))
```

@@

**Key words:** community ecology, epiphyte, fern, functional traits, gametophyte, Moorea, Society Islands, Tahiti.

@@

While many similar adaptations are found in epiphytic seed plants and serve as key reference points [@Benzing1990], the evolution of epiphytism in ferns is more complex owing to their unique life-cycle. 
Ferns produce a free-living gametophyte that is typically smaller than the sporophyte and relies on aquatic motile sperm.
Unlike the sporophyte, the fern gametophyte lacks vascular tissue, stomata, or cuticle, placing it in constant equilibrium with its external environment. 
In parallel with some of the adaptations in the sporophytes, the gametophytes of some epiphytic ferns also exhibit morphological and physiological adaptations to minimize water loss and tolerate water scarcity. 
For example, extreme desiccation tolerance is prevalent in gametophytes of multiple lineages of epiphytic ferns, but is rare in sporophytes [@WatkinsJr.2007c; @Pittermann2013].
In addition, epiphytic fern gametophytes tend to exhibit more complex three-dimensional morphologies that may function to retain water and slow its loss [@WatkinsJr.2007c].
Despite evidence for adaptations to epiphytism in multiple lineages in both sporophytes and gametophytes, no previous studies to our knowledge have analyzed the functional ecology of fern epiphytes in a phylogenetic context including both life stages and representatives of all five Cretaceous epiphytic radiations *sensu* @Schuettpelz2009.

@@

Here, we seek to understand the role of the canopy in fern diversification by investigating traits related to epiphytic growth in fern gametophytes and sporophytes on Moorea, French Polynesia.
Moorea is an isolated tropical oceanic island with high phylogenetic diversity of ferns distributed across a steep elevational gradient, making it ideal for a community phylogenetic analysis of functional traits. 
Specifically, we test the following hypotheses related to epiphytic growth in ferns: 1) Epiphytic ferns have converged upon similar morphologies in both sporophyte and gametophyte phases; 2) Epiphytic fern communities have lower functional and phylogenetic diversity relative to terrestrial communities.

@@

Terrestrial and epiphytic species are partly differentiated in trait space, occupying mostly distinct areas but overlapping in the middle regardless of method used (standard or phylogenetic PCA) (\fig \@ref(fig:pca)*c*, *d*). The first two PC axes explained `r percent(pca_std_cum_var)` and `r percent(pca_phy_cum_var)` of the variation in the trait data when analyzed using standard and phylogenetic PCA, respectively. In the standard PCA, size-related traits tended to load on PC1, with SLA, number of pinnae pairs, and degree of dissection on PC2. In the phylogenetic PCA, SLA was strongly loaded on PC1, while dissection, pinnae, and other size-related traits were mostly loaded on PC2.

@@

In the full-subsets analysis including all climate variables and diversity metrics, growth habit and temperature emerged as the most important variables (\fig \ref{fig:heatmap}). Most of the best-fitting models included growth habit or the interaction of growth habit and a temperature-related variable (\tab S6). Spatial autocorrelation was not detected for any of the diversity metrics (\tab S6).

@@

Despite the apparent constraints on size and the recurrence of non-cordate gametophytes exhibited by epiphytes, we observed greater morphological diversity among epiphytes versus terrestrial species for several traits.
These results do not support the hypothesis that fern epiphytes have reduced functional diversity.
For example, although epiphytes demonstrated less dissected fronds at the species level (\tab \@ref(tab:pics)), they exhibited much greater variation in this trait relative to terrestrial ferns at the community level (\fig S2).
It is likely that the high diversity of this trait reflects different physiological adaptations to various canopy niches. 
For instance, filmy ferns (Hymenophyllaceae), which depend on desiccation tolerance to survive dry conditions, have extremely thin, highly dissected fronds.
Bird’s nest ferns (e.g., *Asplenium nidus* L.), in contrast, avoid desiccation by capturing detritus and water in the bases of their simple, rosette leaves. 
At low-elevation sites on Moorea where these ferns co-occur, filmy ferns frequently grow in more humid portions of trees including near the main trunk, whereas bird’s nest ferns often grow higher in the canopy and on more exposed, drought-prone branches (J. Nitta, pers. obs.). 
Similar predictable patterns of niche partitioning at the local scale have been reported in neotropical epiphtyic ferns as well [@Hietz1998; @Tausz2001]. 
This distribution may reflect alternative water-usage strategies aligned to specific microhabitats in the canopy [desiccation tolerance vs. water capture; @Hietz1998]. 
Such hypotheses need to be carefully tested with additional quantitative investigation, including more finely examining microclimatic variation and species composition across the canopy.

@@

### The primacy of growth habit in determining fern community composition as indicated by modeling
